---
title: "Introduction to R II"
author: "Shaurav Regmi"
date: "April 17, 2018"
output: 
  html_document:
    theme: united
    toc: TRUE
    toc_depth: 4
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

***

```{r message=FALSE, warning=FALSE}
#Loading the packages and data as shown on the previous guide "Introduction to R". 

packages.to.load <- c("plyr", "dplyr" ,"DT", "ggplot2", "plotly", "RODBC")

invisible(lapply(packages.to.load, library, character.only=TRUE))


load("C:/Users/shregmi/Documents/Current Projects/R Training/Example Datasets/Auto.rda")

write.csv(Auto, file="Autodata.csv")

auto.data <- read.csv(file ="C:/Users/shregmi/Documents/Current Projects/R Training/Autodata.csv", header = TRUE)
```


#Manipulating Data

This section will cover a couple ways to shape data into something that is easy to work with.

***

##Using the $ operator

For very simple filters and selection of certain columns from a data frame, you can use the $ operator. Here, we take one column from our auto.data data frame and assign it to a new object. You can use the class() function to see what kind of object it is.
```{r}
JustMPG <- matrix(auto.data$mpg)

class(JustMPG)
```

We can also pull multiple columns pretty easily.
```{r}
MPGandWeight <- data.frame(auto.data$mpg, auto.data$weight)
head(MPGandWeight)
```



***

##dplyr

The package dplyr is a heavily used library for data cleaning and preparation (one of R's main strong suits). Here is a little cheat sheet covering some dplyr and tidyr features: https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

There are 5 main functions or "verbs" used in dplyr: Select, Filter, Mutate, Arrange, and Summarize. Many of these align with the main SQL commands: Select, Where, Group By, etc...

###Select
Let's take a subset of columns from the auto.data dataset.
```{r}
auto.subset <- select(auto.data, name, mpg, cylinders, weight)

head(auto.subset)
```

We can also specify which columns to exclude using the "-" operator.

```{r}
auto.subset2 <- select(auto.data, -horsepower, -year)
head(auto.subset2)
```


###Filter
After selecting your data of interest, you can pass filters in the same way as you would a "where" clause in SQL.

```{r}
only.4cyl <- filter(auto.subset, cylinders==4)
head(only.4cyl)
```

You can add multiple arguments to the filter function.

```{r}
multi.filter <- filter(auto.subset, cylinders %in% c(4,6), weight<3000)
head(multi.filter)
```

We can filter strings as well.

```{r}
filter.name <- filter(auto.subset, cylinders==4, weight<2000, grepl("toyota|volkswagen", name))
head(filter.name)
```


###Mutate
Using mutate, we can create new columns. Additionally, we can chain the previous "verbs" together using the "%>%" operator (called pipe operator). Note that the dplyr package must be loaded in order to use this operator. 

```{r}
add.col <- auto.data %>%
  select(name, mpg, cylinders, weight) %>%
  filter(weight>1800) %>%
  mutate(weightpercyl <- weight/cylinders)

head(add.col)
```


###Arrange
This is equivalent to an "order by" statement in SQL. Here, we sort the previous output by "weightpercyl" in descending order using the "-" operator. 

```{r}
add.col.sorted <- auto.data %>%
  select(name, mpg, cylinders, weight) %>%
  filter(weight>1800) %>%
  mutate(weightpercyl = weight/cylinders) %>%
  arrange(-weightpercyl)

head(add.col.sorted)
```


###Summarize


```{r}
summarized.auto <- auto.data %>%
  select(name, mpg, cylinders, weight) %>%
  filter(weight>1800) %>%
  mutate(weightpercyl = weight/cylinders) %>%
  summarise(avg_mpg = mean(mpg),
            min_weight = min(weight),
            median_weightpercyl = median(weightpercyl))

summarized.auto
```

We can pass a group_by() clause as well.

```{r}
summarized.auto2 <- auto.data %>%
  select(name, mpg, cylinders, weight) %>%
  filter(weight>1800) %>%
  mutate(weightpercyl = weight/cylinders) %>%
  group_by(cylinders) %>%
  summarise(avg_mpg = mean(mpg),
            avg_weight = mean(weight),
            avg_weightpercyl = mean(weightpercyl))

summarized.auto2
```

***





#Visualization

This section will cover a few useful packages and methods for visualization in an R Markdown document. 

***

##Displaying Tables in R Markdown

When using print() or head() functions to show the contents of a data frame, the output looks very ugly. There are a huge number of packages that are specifically for displaying tables in an aesthetically pleasing way. The "DT" or Data Tables package is one of them.

```{r}
datatable(auto.data)
```
This is the default way to display a data frame with this package with no arguments provided except for the object that is to be shown.

There are a huge number of arguments you can add to make your reports look nicer and to add functionality. BONUS: You'll also write the smallest piece of JavaScript code in this as well. 

```{r}
datatable(auto.data, extensions = 'Buttons', options = list(
  dom = 'Bfrtip', 
  buttons= c('copy', 'excel'),
  pageLength = 5,
  initComplete = JS("
    function(settings, json) {
      $(this.api().table().header()).css({
        'background-color': '#232D69',
        'color': '#fff'
      });
    }")
), rownames = FALSE)
```



***


##Charts with Plotly

Cheat sheet for plotly: https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf. 

Another great resource: https://plot.ly/r/ 

```{r}
plot <- plot_ly(x = auto.data$weight, type="histogram")
plot
```

In plotly, you can add multiple sets of traces or bars to the same plot with the use of pipe operators. You can also name each of the traces/bars so that the result is clear. 

```{r}
plot2 <- plot_ly(data =auto.data, alpha = 0.5) %>% #adjusting color transparency 
  add_histogram(x = auto.data$weight[auto.data$origin == 1], name="American") %>%
  add_histogram(x = auto.data$weight[auto.data$origin == 2], name="European") %>%
  add_histogram(x = auto.data$weight[auto.data$origin == 3], name="Japanese") %>%
  layout(barmode = "overlay")

plot2
```

Here is an example of a scatterplot generated by plotly. This has many more features and much more functionality than the plot we first generated with the R Base plot() function. 

```{r}
auto.data$OriginName[which(auto.data$origin == 1)]  = "American"
auto.data$OriginName[which(auto.data$origin == 2)]  = "European"
auto.data$OriginName[which(auto.data$origin == 3)]  = "Japanese"


plot3 <- plot_ly(data = auto.data, x = ~weight, y = ~mpg, color = ~OriginName, type ="scatter", mode="markers", text= ~paste("Car name: ", name ,
                                                                                                                                "</br> Year: ", year,
                                                                                                                                "</br> Cylinders: ", cylinders))

plot3
```
***

##Charts with ggplot2

Cheat sheet for ggplot2: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf 

***

<br>
<br>
<img style="float: right;" src= "https://www.evolenthealth.com/sites/all/themes/evolentcorporate/logo.png", width="200", height = "50">
<br>
<br>
<br>

***
